---
title: "BOO 2025 - Example Analysis"
subtitle: "Script 1: Data exploration - Questions"
date: "`r Sys.Date()`" 
author: 
  Jurgen Zwerwer
output:
  html_document:
    code_download: true
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

***

> In this script you will explore the project data, describing the variables and outlining the data structure. 

***

# Setup

## Using this template

For each of the steps in the BOO project, we have prepared an example script like this one. You can download the R markdown (.Rmd) file to make it easier to complete the steps.

**Exercise 1: Download the .Rmd using the `Code` dropdown menu in the top right corner of this document, and load it into R Studio.**

***

## Clean

It is good practice to clean your environment at the start of every script. That way all the objects you need for the analysis are loaded within the code, and the R markdown document will knit without issues. 

**Exercise 2: Remove all current objects from the R environment:**

<details>
  <summary><strong>Hint</strong></summary>

  `ls()` is a function that returns the names of all objects currently loaded into the session.

</details>

```{r clean}
# 
ls()
```

***

## Load packages

In R, different packages can have functions with the same name. Later loaded packages can therefore mask functions from earlier loaded ones. To avoid confusion when rerunning chunks within a script, we advise loading all required packages at the start. 

As you learned in the DataCamp course, `tidyverse` is a group of packages which includes many basic functions for data wrangling and visualization in R.

**Exercise 3: Load the `tidyverse` package into your environment:**

```{r load-packages}
library(tidyverse)
library(dplyr)
```

You can see from the reported message that loading `tidyverse` masks base R functions, `filter()` and `lag()`. If it is the case that a function you want to use is masked, you can always specify the package you intend to draw from explicitly using the `::` operator (e.g. `stats::filter()`).

***

## Set variables

Setting variables that you will use repeatedly in a script can save time if you need to make changes later. This can include paths to folders where you store files or QC thresholds, for example.

**Exercise 4: Create the following objects in your R environment:**

* `root_dir` - your project folder
* `count_path` - location of the count data
* `metadata_path` - location of the metadata (sample information)

```{r set-variables}
root_dir <- "/Users/jurgenzwerwer/BOO 2025 - BOO CDS Giulia team - Jurgen/Project/QC/Input/"
count_path <- file.path(root_dir, "data", "counts.txt")
metadata_path <- file.path(root_dir, "data", "metadata.txt")
```

***

## Load data

**Exercise 5: Load the project data into your R environment:**

<details>
  <summary><strong>Hint</strong></summary>

  Use the `read_tsv()` function.

</details>

`countData`

* should contain one column `Probe` containing the probe names
* the rest of the columns should represent samples with values containing counts for that particular samples
* this object should have one row for each probe measured by the TempO-Seq

`metaData`

* should have a row for each of the sample columns in `countData`
* should have variables describing the samples in your data

```{r load-data}
Gene_count <- data.table::fread("Input/countData.txt", sep = "\t")
Gene_meta <- data.table::fread("Input/metaData.txt", sep = "\t")
```

***

# Metadata

The `metaData` object contains information on the individual samples in the data.

**Exercise 6: Briefly describe each of the variables in `metaData`:**

```{r metadata}
str(Gene_meta)
# [1] sample_ID: ID of sample       
# [2] sample_type: Treatment or control     
# [3] compound_ID: ID of compound used     
# [4] compound_CAS: CAS number of compound    
# [5] compound_name: name of compound   
# [6] compound_abbr: abbreviation of compound name  
# [7] mean_ID:          
# [8] mean_ID_control: 
# [9] control_type: Type of control    
# [10] conc_ID:  ID for concentration used        
# [11] conc_amt:         
# [12] conc_unit: Unit for concentration      
# [13] conc_max: max concentration used        
# [14] conc_min: min concentration used        
# [15] rep: replicate number of experiment             
# [16] lib_size: total number of reads for sample        
# [17] read_depth:       
# [18] plate_ID: ID of plate        
# [19] plate_well: Which well was used in the plate      
# [20] plate_row: Row of plate       
# [21] plate_col: Column of plate       
# [22] batch: batch number           
# [23] timepoint: time of sample collection       
# [24] cell_type: which type of cell system used       
# [25] compound_class: Class of compounds  
# [26] LOAEL_amt       
# [27] LOAEL_unit: unit of LOAEL      
# [28] PBPK_equivalent: dose extrapolated with PBPK modeling
# [29] PBPK_serum_level: predicted serum concentration based on PBPK modeling
```

***

**Exercise 7: What test system is this data profiled from? Give a brief description.**

<details>
  <summary><strong>Hint</strong></summary>

  Try looking at the `cell_type` variable.

</details>

```{r test-system}
# Primary Human Liver Cells (PHHs) simulate the function of real human liver cells, they are used in metabolism, DDI and toxicity studies. 

```

***

**Exercise 8: How many treatment and control samples are there in the project data?**

```{r sample-type}
# 24 controls, 42 treatments
```

***

**Exercise 9: Describe the range of experimental conditions which the samples have been exposed to:**

<details>
  <summary><strong>Hint</strong></summary>

  Don't forget to describe the compounds and the concentration ranges.

</details>

```{r experimental-conditions}
# For the control PHHs were exposed to either DMSOLow or DMSOHigh. The PHHs with sample_type = treatment were treated with either Chloroacetamide or TMA in concentrations 0.03, 0.30, 1.00, 3.10, 10.00, 32.00 and 180 µM.

```

***

**Exercise 10: How many replicates are there for each experimental condition?**

```{r replicate-number}
# 3

```

***

**Exercise 11: Are the compounds on different plates? If so, how might this affect the analysis and results? Has the experimental design been adjusted to compensate for this?**

```{r plates}
# Treatment and their controls are on the same plate, PHHs exposed to Chloroacetamide and TMA were on different plates. The difference between plates can cause variation in the data. I think so, but don't know how.

```

***

**Exercise 12: Why do some samples not have data on their toxicity class?**

<details>
  <summary><strong>Hint</strong></summary>

  You can filter data using the `is.na()` function.

</details>

```{r na-class}
# Because they are controls treated only with DMSO.

```

***

**Exercise 13: How many hours after exposure are the samples collected?**

```{r timepoint}
# 24 hours
```

***

**Exercise 14: Check that each sample has a unique identifier:**

```{r unique-id}
# This is true; under the column sample_ID

```

***

# Count data

The count data has information on raw counts from the probes measured during TempO-Seq (rows) in each of the samples (columns)

Many of the functions we will use in this project assume that `countData` row names are the gene or probe names. 

**Exercise 15: Change the variable `Probe` to the row names:**

<details>
  <summary><strong>Hint</strong></summary>

  Use the `column_to_rownames()` function.

</details>

```{r column-rownames}
column_to_rownames(Gene_count, "Probe") 
```

***

**Exercise 16: Which column in the metadata can be used to subset the count data? Use it to select only counts from the control samples:**

<details>
  <summary><strong>Hint</strong></summary>

  Make sure you don't replace `countData` with only control samples when answering this question! :)

</details>

```{r subset-controls}
Gene_count <- as.data.frame(Gene_count)

control_samples <- Gene_meta$sample_ID[Gene_meta$sample_type == "Control"]

control_counts <- Gene_count[, control_samples]

control_counts
```
***

**Exercise 17: Check that the columns of the count data and rows of the metadata are in the same order.**

<details>
  <summary><strong>Hint</strong></summary>

  The different functionalities of `==` and `%in%` may help here.

</details>

```{r check-order}
Gene_count <- Gene_count[, Gene_meta$sample_ID]
all(colnames(Gene_count) == Gene_meta$sample_ID)
```

***

# Advanced question

This question provides a bit of a (hopefully fun) challenge, but you can also skip it, save your data, and move on to the probe QC if you prefer.

**Exercise 18: Make a plot of the plate layout using `ggplot`?**

<details>
  <summary><strong>Hint</strong></summary>

  Use `geom_tile()`

</details>

```{r plate-layout}


```

***

# Save

Make sure you save any changed objects at the end of your script for downstream analysis.

**Exercise 19: Save your `countData` and `metaData` as `.Rdata` objects**

<details>
  <summary><strong>Hint</strong></summary>

  Don't forget to use `root_dir`, so if you ever move your project your code can be easily updated.

</details>

```{r save-data}
# 

```

***

# Session Info

Printing `session_info()` at the bottom of your scripts ensures your work is reproducible. It does this by keeping a record of the package versions used and the environment. 

**Exercise 20: Print your session info here:**

```{r session-info}
# 

```

***

**Exercise 21: Knit this script to produce either a `PDF` or `HTML`file and push it to your GitHub repository.**

That is the end of the Data Exploration. Example answers will be available from the `BOO_template` GitHub on Tuesday. 

Next, please move on to the probe QC using `02_ProbeQC_Outline.Rmd`.

***

