---
title: "BOO 2025 - Example Analysis"
subtitle: "Script 4: PCA - Questions"
date: "`r Sys.Date()`" 
author: 
  Jurgen Zwerwer
output:
  html_document:
    code_download: true
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

***

> In this script, you will perform principal component analysis (PCA) to further explore patterns in the project data.

***

# Setup

## Clean

As before, we perform several setup steps at the start of the script to ensure our work is reproducible and clear. 

**Exercise 1: Download the R markdown template, clean your environment, and set the following variables:**

* `root_dir` - project folder
* `cpm_path` - location of the `cpmData` object
* `metadata_path` - location of the `metaData` object

```{r clean}
root_dir <- "~/Documents/BOO 2025 Code/Project/QC/Output/"
cpm_path <- "cpmData_QC.Rdata"
count_path <- "countData_QC.Rdata"
metadata_path <- "metaData_QC.Rdata" 
```

***

## Packages

Two new packages are introduced in this script:

* `ComplexHeatmap` allows the drawing and annotation of heatmaps in R
* `circlize` allows for the drawing of circular plots, but is also used by `ComplexHeatmap` for colour functions like `colorRamp2()`

**Exercise 2: Load these packages alongside `tidyverse` into your R environment:**

```{r load-packages, warning=F, message=F}
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
```

***

## Load data

**Exercise 3: Load the CPM and metadata into your R environment:**

```{r load-data, warning=FALSE, message=FALSE}
load(paste0(root_dir, cpm_path))
load(paste0(root_dir, count_path))
load(paste0(root_dir, metadata_path))
```

***

# PCA

## Calculate

In high dimensional data (such as this data with around 10,000 genes), principal components (PCs) can be calculated and used to explore patterns. PCs can be thought of as a new axis in the data that captures the most variance possible in the fewest variables. 

**Exercise 4: Use the `prcomp()` function to calculate PCs of the `cpmData`:**

```{r pca-calc}
# The PCs that capture CPM data variance are calculated
pcs <- prcomp(t(cpmData))
```

Tolerance (or `tol`) can be adjusted to create more or fewer PCs, where a lower tolerance generates a higher number. If this argument is not set, the PCs calculated will capture the full variability of the CPM data.

***

## Variance explained

**Exercise 5: Use the output of `prcomp()` to explore your PCs and calculate the variance in CPM values that they explain:**

<details>
  <summary><strong>Hint</strong></summary>

  Variance explained is the SD squared divided by the sum of the variance for all PCs. 

</details>

```{r pc-summ}
# First the PCs are summarized
summary(pcs)
# Then the explained variance is calculated
var_explained =
  data.frame(PC = 1:nrow(Gene_meta),
             var_explained = round(pcs$sdev^2 / sum(pcs$sdev^2), 3))

var_explained
```

***

## Screeplot

A screeplot can be used to visualize how each subsequent PC captures less and less of the total variance in the data.

**Exercise 6: Plot the top 20 calculated PCs against the variance they explain to generate a screeplot:**

```{r screeplot}
var_explained %>% filter(PC <= 20) %>%

ggplot(aes(x=PC, y = var_explained)) + 
  geom_line(color = "grey40") +
  geom_point(color = "grey40", fill = 'lightblue', shape = 21, size = 3) + 
  scale_x_continuous(breaks = c(seq(1,20))) +
  xlab("Principal Component") +
  ylab("Proportion of variance explained") +
ggtitle("Screeplot of the first 20 PCs") +  theme_bw() +
  theme(axis.text.x = element_text(size = 8))
```

***

## Relation to known variables

By investigating how PCs correlate with known variables, we can assess how much each factor impacts expression. 

**Exercise 7: Add the PCs that explain more than 1% variance in CPM values to the metaData for further investigation:**

```{r add-pc}
# The first 10 PCs are added to the metadata
Gene_meta <- cbind(Gene_meta, pcs$x[,1:10])
```

***

Correlations between known factors and PCs can be calculated using the `cor()` function, which was used for the replicate correlation in the sample QC.

**Exercise 8: Generate a matrix of correlations between PCs explaining over 1% of CPM variance and known factors**

<details>
  <summary><strong>Hint</strong></summary>

  Variables that have a standard deviation above 0 will have a correlation of NA, so you may want to remove these.

</details>

```{r plot-vars, warning=F, message=F}
# The SD of each column was calculated in Gene_meta
plot_vars <- apply(Gene_meta, 2, function(x) sd(as.numeric(factor(x)), na.rm=T))
# Column names that vary are saved (SD is not NA or 0)
plot_vars <- names(plot_vars[!plot_vars %in% c(NA, 0)])
# PC column names are then removed
plot_vars <- plot_vars[!grepl("PC", plot_vars)]

plot_vars
```
```{r}
# We subset Gene_meta to include only these columns
heatmap_df <- Gene_meta %>% select(any_of(plot_vars))

# All variables are converted to numeric values
heatmap_df <- apply(heatmap_df, 2, function(x) as.numeric(factor(x)))

# Then we calculate correlations between the first 10 PCs and these numeric variables scaled
cxy <- round(cor(pcs$x[,1:10], scale(heatmap_df), 
                 use = "pairwise.complete.obs"), 2)

as.data.frame(cxy)
```

***

Such a correlation matrix can be visualized using a heatmap.

**Exercise 9: Create a heatmap of correlations between known factors and the selected PCs:**

<details>
  <summary><strong>Hint</strong></summary>

  `colorRamp2` can be used to generate a custom colour palette.

</details>

```{r heatmap}
# To visulize how correlated the variables are, a color scale from -1 to 1 for the heatmap is created.
col_fun <- colorRamp2(c(-1, 0, 1), c("#008080", "white", "#b3002d"))

# Then a heatmap is created
Heatmap(
  t(cxy),         
  # A color palette is added
  col = col_fun,  
  border = 'grey5',
  # Then the variables are clustered but not the PCs
  cluster_columns = FALSE,            
  show_row_dend = TRUE,             
  show_column_dend = FALSE,    
  # The name of legend is set
  name = "Corr",      
  # Then the text is formatted
  row_names_gp = gpar(fontsize = 8), 
  column_names_gp = gpar(fontsize = 8), 
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(x, y, width, height, 
              gp = gpar(col = "white", lwd = 1, fill = NA))
    # The correlation is printed if it is above 0.4
    grid.text(ifelse(abs(t(cxy)[i,j]) > 0.4,
                     sprintf("%.2f", round(t(cxy)[i, j], 2)),
                     ""), 
              x, y, gp = gpar(fontsize = 8, col = "white"))
  }
)
```
PC1 and PC3 show high correlations. PC1 with plate_ID and PC3 with batch. PC4 also has high correlations but in other variables. PC1 was compared to both PC3 and PC4 below.

## PCA plot

**Exercise 10: Make a plot of two important PCs against each other, coloured by a relevant factor:**

<details>
  <summary><strong>Hint</strong></summary>

  You can use different shapes to visualize multiple factors in the same plot.

</details>

```{r pca-plot}
Gene_meta %>% 
  # PC1 is plotted against PC3
  ggplot(aes(x = PC1, y = PC3, 
             #  The data points are coloured by high-tox and low-tox and with different shapes for each plate
             color = compound_class, shape=plate_ID)) +
  geom_point(size = 2) +
  # Then labs are added to the x and y axis
  labs(x = paste0("PC1 (", round(100*var_explained[2,2], 2), "%)"), 
       y = paste0("PC3 (", round(100*var_explained[3,2], 2), "%)"), 
       color = "Class", shape = "Plate") +
  ggtitle("PCA plot") +
  theme_bw()
```
The high- and low-tox compounds form distinct groups suggesting that they have different effects. A plate effect can be observed as the triangles and dots are clearly separated.
```{r}
#The same code was used as before, only with PC4 on the y-axis instead of PC3
Gene_meta %>% 
  ggplot(aes(x = PC1, y = PC4, 
             color = compound_class, shape=plate_ID)) +
  geom_point(size = 2) +
  labs(x = paste0("PC1 (", round(100*var_explained[2,2], 2), "%)"), 
       y = paste0("PC4 (", round(100*var_explained[3,2], 2), "%)"), 
       color = "Class", shape = "Plate") +
  ggtitle("PCA plot") +
  theme_bw()
```
The same pattern can be seen compared to the PCA plot of PC1 vs PC3.
***

# Advanced questions

Sometimes a PCA plot can highlight important clusters in your data. Gene loadings can be used to assess which gene's expression is driving these clusters.

**Exercise 11: Investigate a pattern in your data and identify what genes are responsible for it:**

```{r advanced-pca}
#

```

***

# Session Info

**Exercise 12: Print your session info at the end of the script, knit the R markdown document, and push it to GitHub:**

```{r session-info}
#
sessionInfo()
```

***

That is the end of the entire QC pipeline. Example answers will be available from the `BOO_template` GitHub on Tuesday. 

***
